Unifying intraspecific and interspecific variation in tropical tree mortality

### 2016-07-04 ms/ms.Rnw
Move methods to supplementary

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[a,1]{James S Camac}
\author[a,2]{Mark Westoby} 
\author[a,3]{Daniel S Falster}

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\affil[a]{Department of Biological Sciences, Macquarie University NSW 2109, Australia}

@@

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{J.C. designed and performed the analysis; D.F. conceived the idea, supervized the development of this work; M.W. provided biological insight. All authors contributed to the writing of this manuscript.}

@@

\begin{abstract}
Individual death is a key process driving population dynamics, nutrient cycling, and evolution within plant communities. While past research has identified factors across a variety of scales influencing tree mortality -- including an individual's past growth, species-level functional traits, and temporal plot-scale effects; these different factors are yet to be integrated within a predictive framework. Here we apply classic techniques from survival analysis within a hierarchical Bayesian and cross-validation framework to derive instantaneous mortality functions for tropical rain forest trees at Barro Colorado Island (BCI) Panama, integrating individual-level, species-level, and plot-level effects. In reaching this point, we show how mortality models including separate growth-dependent and growth-independent components, using stem diameter growth rather than basal-area growth, and attributing the effects of wood density to growth-independent death outperform alternative formulations. Moreover, we show that the effects of wood density -- a prominent trait known to influence mortality -- capture only a portion of total species-level differences; and that the year-to-year variation in mortality was large.  Our analysis also suggests that growth-dependent processes -- and thus competition -- predominate as the biggest driver of mortality at BCI. Combined, this study provides a framework for predicting individual-level mortality in highly diverse tropical forests, but also highlights how little we now about the causes of species-level and temporal plot-scale effects, needed to effectively predict tree mortality.
\end{abstract}

@@

However, due to a paucity of data on plant mortality, most dynamic vegetation models have implemented mortality algorithms with overly simplistic assumptions \cite{Hawkes:2000ib, Wyckoff:2000un, McDowell:2011dr}. One of the strongest assumptions made is that both stochastic hazards and the growth-mortality relationship are constant across species \cite{Pacala:1996vm, Hawkes:2000ib, Wyckoff:2000un}. That is, all species have the same random chance of death and exhibit the same tolerance to low growth. This is despite growing empirical evidence highlighting that mortality rates vary considerably between species, even within the same community \cite{Kobe:1997vy, Wyckoff:2002ul, Kunstler:2005bn, Vieilledent:2010fv}. Furthermore, because no systematic analysis has been conducted to determine what growth measure best predicts plant mortality, the growth-mortality relationship is often parameterized using a variety growth measures; the two most common being rates of DBH or basal area growth.

@@

In this study, we reconcile both intraspecific and interspecific theory by developing a unified model that incorporates both individual growth and the species level trait, wood density. We extend existing individual-based approaches \cite[e.g.][]{Kobe:1995tw} by simultaneously estimating both baseline hazards and the growth-mortality relationship within a single model (hereafter referred to as growth-dependent hazard). We then compare the predictive performance of this model to other traditionally used mortality models. We parameterize these models using 15-years of individual growth and mortality data from 203 tropical rainforest species encompassing \Sexpr{n_inds} individuals. Using these models, coupled with 10-fold cross validation, we ask several questions: 1) Does simultaneously estimating baseline and growth-dependent hazards improve predictive capacity relative to baseline only or growth-dependent only models? 2) is DBH growth or basal area growth more predictive of tropical tree mortality and does it's predictive capacity vary with model type? 3) Does allowing both hazards to vary by species improve predictive capacity? 4) Does wood density predominately affect baseline hazards, or does it affect growth-dependent hazards or both?

@@

In classic survival analysis, the probability that an individual dies between $t_2$ and $t_3$, $p_{t_2\rightarrow t_3}$ is defined by
\begin{equation} \label{eq:mortality_model}
p_{i,t_2\rightarrow t_3} = 1 - \exp\left(- \int_{t_2}^{t_3} \lambda_i(t) {\rm d}t\right),
\end{equation}
where $\lambda_i(t)$ is the instantaneous hazard (mortality) rate of death per unit time \cite{Kleinbaum:2012tn} and $\int_{t_2}^{t_3} \lambda_i(t) {\rm d}t$ is the cumulative hazard between $t_2$ and $t_3$. The challenge then is to identify suitable forms of the instantaneous hazard (mortality) rate function.

We examine the performance of three functional forms of $\lambda_i(t)$ to predict observed patterns of tropical tree mortality:
\begin{align}
\lambda_i &= \overbrace{\gamma}^{\text{baseline hazard}} \label{base_haz}\\
\lambda_i &= \overbrace{\alpha \, \exp\left(-\beta \, X_i \right)}^{\text{Growth-dependent hazard}} \label{growth_haz}\\
\lambda_i &= \overbrace{\alpha \, \exp\left(-\beta \, X_i \right)}^{\text{Growth-dependent hazard}} +
\overbrace{\gamma.}^{\text{Baseline hazard}} \label{add_haz}
\end{align}
The simplest model (eq. \ref{base_haz}), assumes a single baseline hazard, which is constant irrespective of a plant's growth rate. The second model (eq. \ref{growth_haz}) describes a growth-dependent hazard which assumes plants can only die due to a growth related cause. In this model, we make an explicit assumption that the growth-mortality relationship is captured by a negative exponential curve whereby the probability of death is high at low growth rates and as growth increases the probability of dying decays at an exponential rate until an asymptote at zero is reached. This negative exponential relationship has been used in other studies to model growth-dependent mortality \cite[e.g.][]{Kobe:1995tw} and broadly reflects patterns observed in non-parametric growth-mortality models \cite{Wyckoff:2000un, Kunstler:2005bn, Vieilledent:2010fv}. The third model (eq. \ref{add_haz}) -- potentially the most biologically realistic -- is the summation of the two previous models. Specifically, this model allows $\lambda$ to decrease exponentially with increasing growth, but, unlike a standard negative exponential, will asymptote at some baseline hazard instead of zero. The advantage of this model is that it effectively accounts for two independent causes of mortality: death due to growth and death due to chance (i.e. stochastic events such as wind throw and fire). The parameters of eq. \ref{add_haz} are also biologically interpretable: $\alpha$ represents the intercept and defines the instantaneous mortality hazard $\lambda$ at low annual growth rate. $\beta$ represents the exponential decay rate or the sensitivity of hazard rates to change with growth rate; and $\gamma$, is the asymptote and defines the baseline hazard. $\alpha$ and $\beta$ are intrinsically linked to a plants growth rate, and as such are growth-dependent parameters. By contrast, $\gamma$ is a growth-independent parameter that accounts for stochastic events (e.g. windfall, fire) that can kill a plant irrespective of its growth rate.

@@

@@

Field measurements of diameter inevitably include observation error, errors that may ultimately bias conclusions. In our dataset, \Sexpr{proportion_negative}\% of observed growth rates were negative. While there are physiological processes (e.g. water depletion and bark drop) that can lead to trunk shrinkages overtime \cite{Pastur:2007kl}, it is biologically impossible for growth (i.e. the creation of new biomass) to be negative \cite{Clark:2007ve}. To ensure our model was not biased by these implausible values we estimated \textit{``true growth''}(i.e. observed growth minus observation error) by first using another BCI dataset collected explicitly to examine observation errors \cite{Condit:2012nz}. This dataset consisted of \Sexpr{n_obs_errors} randomly selected trees that had their diameter measured twice within a 30 day period \cite{Condit:2012nz}. Assuming no growth occurred over this period, any discrepancy in measurements can be attributable to observation error. The standard deviation of these discrepancies was estimated to be \Sexpr{obs_error_sigma}, meaning that 95\% of the measurement errors were within $\pm$ 1.47 cm of the true diameter. Having an estimate of observation error, we then used Bayesian point estimation to maximize the joint posterior of the \textit{``true''} initial and final diameter, and subsequently, \textit{``true growth''} for each individual. We did this by first modelling the observed initial ($D_{i,1}$) and final diameter ($D_{i,2}$) of each individual, $i$, as random realisations from a normal distribution centerd on their respective true estimates (i.e. $\widehat{D_{i,1}}$ and $\widehat{D_{i,2}}$), with the standard deviation fixed at \Sexpr{obs_error_sigma} (i.e. the observed measurement error):

\begin{equation}\label{eq:true_growth_obs} 
\begin{aligned}
D_{i,1} &\sim \mathcal{N}(\widehat{D}_{i,1}, 0.75) \\
D_{i,2} &\sim \mathcal{N}(\widehat{D}_{i,2}, 0.75)
\end{aligned} 
\end{equation}

The initial diameter for each individual $i$, $\widehat{D}_{i,1}$, was sampled as a random realisation from a log-normal distribution:
\begin{equation} \label{eq:true_dbh1}
\widehat{D}_{i,1} \sim \log \mathcal{N}(\mu_{D_1}, \sigma_{D_1}),
\end{equation}
where $\mu_{D_1}$ is the expected mean of log initial diameter and $\sigma_{D_1}$ defines that variation around that mean. $\mu_{D_1}$ was estimated using a normal prior $\mathcal{N}(0, 2.5)$, which represents our prior expectation that the mean initial diameter would be greater than zero but less than 150 cm. $\sigma_{D_1}$ was estimated using a weakly informative half Cauchy prior centerd on zero and a scale parameter of 2.5.

The final diameter measurement for each individual $i$, $D_{i,2}$ was estimated as:
\begin{equation}\label{eq:true_dbh2}
\widehat{D}_{i,2} = \widehat{D}_{i,1} + \Delta D_{i,2} \, (t_2-t_1),
\end{equation}
where $\Delta D_{i,2}$ represents the annual true increase in diameter (cm) and $(t_2-t_1)$ represents the number of days between initial and final measurements divided by 365.25. We modelled $\Delta D_{i,2}$ as a random realisation from a log-normal distribution centerd on zero and a standard deviation of 1. This prior conveyed our expectation that annual diameter growth could not be negative and is unlikely to be greater than 7 cm/yr.

Using this model we then used \texttt{rstan}'s \textit{optimizing} function to maximize the joint posterior and extract point estimates of true initial diameter, $\widehat{D}_{i,1}$, true final diameter, $\widehat{D}_{i,2}$, and true annual diameter growth $\Delta D_{i,2}$. We then used annual diameter growth, $\Delta D_{i,2}$, and annual basal area growth: $\frac{(\pi \, 0.25 \,\widehat{D}_{i,2}) - (\pi \, 0.25 \, \widehat{D}_{i,1})}{(t_2-t_1)}$ in the subsequent mortality analysis.

\subsection*{Model fitting}
Our observations are for survival outcomes $S_{i, t_2\rightarrow t_3}$ (0 = alive, 1 = died) for each individual $i$ between censuses at time $t_2$ and $t_3$. We model these observations as random realizations from a Bernoulli distribution
\begin{equation}
S_{i,t_2\rightarrow t_3} \sim \textrm{Bernoulli}(p_{i,t_2\rightarrow t_3}),
\end{equation}
where $p_{i,t_2\rightarrow t_3}$ is the probability individual $i$ dies in the interval $t_2$--$t_3$ *(eq. \ref{eq:mortality_model}). As we only have a point estimate of $\lambda_i(t)$ at $t=t_2$, the cumulative hazard from  $t_2$--$t_3$ is approximated as
\begin{equation}
  \int_{t_2}^{t_3} \lambda_i(t) {\rm d}t \approx \lambda_i(t_2) (t_3-t_2).
\end{equation}

As we have multiple hypothesized functional forms of $\lambda_i$ (e.g. equations \ref{base_haz}\textendash\ref{add_haz}), we describe here the fitting procedure of our most complicated model, of which all others may be considered special cases. The instantaneous rate of mortality is
\begin{equation}
\lambda_i = \left(\overbrace{\alpha_{s[i]} \, \exp\left(-\beta_{s[i]} \, X_i \right)}^{\text{Growth dependent}} +
\overbrace{\gamma_{s[i]}}^{\text{Growth independent}}\right) \times \overbrace{\delta_{t[i]}}^{\text{Census effect}}
\end{equation}  
In this model, $\alpha_s$, $\beta_s$ and $\gamma_s$ are allowed to vary by species as a function of both wood density (measured at species level) and a species random effects, e.g.
\begin{equation}\label{eq:submodels}
\alpha_s = \alpha_{0,s} \, \left(\frac{\rho}{\rho_c}\right)^{\alpha_{1}},
\end{equation}
with similar formulations for $\beta_s$ and $\gamma_s$. Here $\alpha_{1}$ captures the effect of wood density $\rho$ on $\alpha_s$, while $\alpha_{0,s}$ captures any other species-level residual error not explained by wood density for species $s$ within which individual $i$ belongs. The form of sub-model in eq. \ref{eq:submodels} ensures that parameters $\alpha_s$, $\beta_s$ and $\gamma_s$ remain positive. On a log scale this model equates to an additive linear model , e.g. $\log \alpha_s = \log \alpha_{0,s} + (\alpha_{1} - \log \rho_c) \, \log \rho$.

We standardized wood density via the transformation: $\frac{\rho}{\rho_c}$, where $\rho_c=0.6$ g cm${^-3}$. This operation is equivalent to centering a log-transformed $\rho$ around $\rho_c$ and enables mean hyper-parameters  to be interpreted as the value for a species with average wood density of 0.6 cm${^-3}$. Similarly, we also centered $X_i$ at the lower 5\% quantile for both diameter increment and area growth (0.172 and 0.338, respectively), meaning $\alpha_s$ should be interpreted as the hazard rate when growth rate was very low.

We modelled species random effects as random realisations from log-normal distributions:
\begin{equation}
\alpha_{0,s} \sim \log \mathcal{N}\left(\mu_{\log \alpha_0}, \sigma_{\log \alpha_0}\right),
\end{equation}
with similar formulations for $\beta_{0,s}$ and $\gamma_{0,s}$.

All three mean hyperparameters, $\mu_{\log \alpha_0}$, the average log species residual error at zero growth, $\mu_{\log \beta_0}$, the average species error on the exponential decay rate, and $\mu_{\log \gamma_0}$, the average baseline log species residual error, were estimated using normal distributions centered on zero with a standard deviation of 2.5.
\begin{equation}
\mu_{\log \alpha_{0}},\mu_{\log \beta_0},\mu_{\log \gamma_0} \sim \mathcal{N}\left(0, 2.5\right).\\
\end{equation}

Standard deviations associated with species random effects (i.e. $\sigma_{a0}$, $\sigma_{b0}$, $\sigma_{c0}$), were estimated using positive half Cauchy priors centered on zero with a scale of 2.5.
\begin{equation}
\sigma_{a0},\sigma_{b0},\sigma_{c0} \sim \textrm{Cauchy+}\left(0, 2.5\right).\\
\end{equation}

For models that did not include species random effects, single intercepts were estimated for each parameter using log normal distributions centered on zero with standard deviations of 2.5.

We estimated wood density effects (i.e. $\alpha_1$, $\beta_1$ and $\gamma_1$) using normal priors centered on zero and a standard deviation of 2.5:
\begin{equation} \label{eq:rho_effects}
\alpha_1, \beta_1, \gamma_1 \sim \mathcal{N}\left(0, 2.5\right).\\
\end{equation}

We accounted for possible variation in mortality rates across different census periods by allowing $\lambda_i$ to be scaled by $\delta_t$ for each census $t$. We estimated $\delta_t$ as a random realization from a log-normal distribution centered on zero and a variance parameter $\sigma_{\delta}$ defines the variation in census effects:

\begin{equation} \label{eq:census_effect}
\delta_t \sim \log \mathcal{N}\left(0, \sigma_{\delta}\right)\\
\sigma_{\delta} \sim \textrm{Cauchy+}\left(0, 2.5\right).
\end{equation}

For each model we used Bayesian inference and fitted models in R \Sexpr{paste(version$major, version$minor, sep='.')} using package \texttt{rstan} \Sexpr{packageVersion("rstan")}\cite{StanDevelopmentTeam:2015uz}. We executed three independent chains and in all cases modelled parameters converged within 2000 iterations. Convergence was assessed through both visual inspection of chains and reference to the Brooks-Gelman-Rubin convergence diagnostic \cite{Gelman:1992ts, Brooks:1998ju}. After discarding the first 2000 iterations as ‘burn in’, a further 2000 iterations were taken from the joint posterior.

>>>>

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[a,1]{James S Camac}
\author[a]{Mark Westoby} 
\author[b]{S Joseph Wright}
\author[a]{Daniel S Falster}

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\affil[a]{Department of Biological Sciences, Macquarie University NSW 2109, Australia}
\affil[b]{Smithsonian Tropical Research Institute, Apartado 0843–03092, Balboa, Panama}

@@

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{J.C. designed and performed the analysis; D.F. conceived the idea, supervized the development of this work; S.J.W \& M.W. provided biological insight. All authors contributed to the writing of this manuscript.}

@@

\begin{abstract}
Death is an integral process driving population dynamics, nutrient cycling, and evolution within plant communities. While past research has identified factors influencing tree mortality -- including an individual's past growth, species-level functional traits, and temporal plot-scale effects; these are yet to be integrated within a general predictive framework. Here we estimate instantenous mortality rates of 203 tropical rainforest tree species at Barro Colorado Island (BCI), Panama, by combining classic survival analysis techniques with a cross-validated hierarchical Bayesian framework that integrates individual-level, species-level, and plot-level effects. We show that greatest predictive capacity is achieved when when models simultaneously estimate growth-dependent and growth-independent hazards, use stem diameter growth rather than basal-area growth and account for wood density effects on growth-independent terms. Moreover, we show that the effects of wood density -- a prominent trait known to influence mortality -- capture only a portion of total species-level differences; and that the year-to-year variation in mortality was large.  Our analysis also suggests that growth-dependent processes -- and thus competition -- predominate as the largest driver of mortality at BCI. Combined, this study provides a framework for predicting individual-level mortality in highly diverse tropical forests, but also highlights how little we now about the causes of species-level and temporal plot-scale effects, needed to effectively predict tree mortality.
\end{abstract}

@@

However, due to a paucity of data on plant mortality, most dynamic vegetation models have implemented mortality algorithms with overly simplistic assumptions \cite{Hawkes:2000ib, Wyckoff:2000un, McDowell:2011dr}. One of the strongest assumptions made is that both stochastic hazards and the growth-mortality relationship are constant across species \cite{Pacala:1996vm, Hawkes:2000ib, Wyckoff:2000un}. That is, all species have the same random chance of death and exhibit the same tolerance to low growth. This is despite growing empirical evidence highlighting that mortality rates vary considerably between species, even within the same community \cite{Kobe:1997vy, Wyckoff:2002ul, Kunstler:2005bn, Vieilledent:2010fv}. Furthermore, because no systematic analysis has been conducted to determine what growth measure best predicts plant mortality, the growth-mortality relationship is often parameterized using a variety growth measures; two of the most common being rates of DBH or basal area growth.

@@

In this study, we reconcile both intraspecific and interspecific mortality theory by combining classic survival analysis techniques with a cross-validated hierarchical Bayesian framework. Specifically, our framework  instantaneous mortality rates for 203 tropical rainforest trees at Barro Colorado Island (BCI) by integrating individual-level, species-level, and plot-level effects and simultaneously estimating growth-dependent and growth-independent mortality processes. Using this framework we 
identify the most parsimonious model of tropical tree instantaneous mortality rates. We also ask: 1) which  growth measure (i.e. diameter increment growth or area growth)best predicts growth-dependent mortality; 2) Does including wood density improve model predictive capacity, and if so, does it predominately affect baseline hazard, growth-dependent hazard or both? 3) What is the proportion of species variability explained by wood density?

@@

In classic survival analysis, the probability that an individual dies between $t_2$ and $t_3$, $p_{t_2\rightarrow t_3}$ is defined by
\begin{equation} \label{eq:mortality_model}
p_{i,t_2\rightarrow t_3} = 1 - \exp\left(- \int_{t_2}^{t_3} \lambda_i(t) {\rm d}t\right),
\end{equation}
where $\lambda_i(t)$ is the instantaneous mortality rate of death per unit time \cite{Kleinbaum:2012tn} and $\int_{t_2}^{t_3} \lambda_i(t) {\rm d}t$ is the cumulative hazard between $t_2$ and $t_3$. The challenge then is to identify suitable forms of the instantaneous hazard (mortality) rate function.

We examine the performance of three functional forms of $\lambda_i(t)$ to predict observed patterns of tropical tree mortality:
\begin{align}
\lambda_i &= \overbrace{\gamma}^{\text{baseline hazard}} \label{base_haz}\\
\lambda_i &= \overbrace{\alpha \, \exp\left(-\beta \, X_i \right)}^{\text{Growth-dependent hazard}} \label{growth_haz}\\
\lambda_i &= \overbrace{\alpha \, \exp\left(-\beta \, X_i \right)}^{\text{Growth-dependent hazard}} +
\overbrace{\gamma.}^{\text{Baseline hazard}} \label{add_haz}
\end{align}
The simplest model (eq. \ref{base_haz}), assumes a single baseline mortality rate, which is constant irrespective of a plant's growth rate. The second model (eq. \ref{growth_haz}) describes a growth-dependent mortality rate which assumes plants can only die due to a growth related cause. In this model, we make an explicit assumption that the growth-mortality relationship is captured by a negative exponential curve whereby the probability of death is high at low growth rates and as growth increases the probability of dying decays at an exponential rate until an asymptote at zero is reached. This negative exponential relationship has been used in other studies to model growth-dependent mortality \cite[e.g.][]{Kobe:1995tw} and broadly reflects patterns observed in non-parametric growth-mortality models \cite{Wyckoff:2000un, Kunstler:2005bn, Vieilledent:2010fv}. The third model (eq. \ref{add_haz}) -- potentially the most biologically realistic -- is the summation of the two previous models. Specifically, this model allows $\lambda$ to decrease exponentially with increasing growth, but, unlike eq. \ref{growth_haz}, will asymptote at some baseline rate instead of zero. The advantage of this model is that it effectively accounts for two independent causes of mortality: death due to growth and death due to chance (i.e. stochastic events such as wind throw and fire). The parameters of eq. \ref{add_haz} are also biologically interpretable: $\alpha$ represents the intercept and defines the instantaneous mortality rate $\lambda$ at low annual growth rate. $\beta$ represents the exponential decay rate or the sensitivity of mortality rates to change with growth rate; and $\gamma$, is the asymptote that defines the baseline hazard. $\alpha$ and $\beta$ are intrinsically linked to a plants growth rate, and as such are growth-dependent parameters. By contrast, $\gamma$ is a growth-independent parameter that accounts for stochastic events (e.g. windfall, fire) that can kill a plant irrespective of its growth rate.

@@

\begin{figure}[ht]
\includegraphics{figures/fig5.pdf}[ht]
\caption{Predicted instantaneous mortality rates its derivative annual mortality probability. A,C) Effect of high (0.8) and low (0.3) wood density on growth-mortality relationship. B,D) Predicted growth\textendash mortality response curves for each species. Errors are $\pm$ 95\% BCI.}
\label{fig5}
\end{figure}

\begin{figure}[ht]
\includegraphics{figures/fig6.pdf}[ht]
\caption{Predicted instantaneous mortality rates its derivative annual mortality probability. A,C) Effect of high (0.8) and low (0.3) wood density on growth-mortality relationship. B,D) Predicted growth\textendash mortality response curves for each species. Errors are $\pm$ 95\% BCI.}
\label{fig6}
\end{figure}

@@

Field measurements of diameter inevitably include observation error, errors that may ultimately bias conclusions. In our dataset, \Sexpr{proportion_negative}\% of observed growth rates were negative. While there are physiological processes (e.g. water depletion and bark drop) that can lead to trunk shrinkages overtime \cite{Pastur:2007kl}, it is biologically impossible for growth (i.e. the creation of new biomass) to be negative \cite{Clark:2007ve}. To ensure our model was not biased by these implausible values we estimated \textit{``true growth''}(i.e. observed growth minus observation error) by first using another BCI dataset collected explicitly to examine observation errors \cite{Condit:2012nz}. This dataset consisted of \Sexpr{n_obs_errors} randomly selected trees that had their diameter measured twice within a 30 day period \cite{Condit:2012nz}. Assuming no growth occurred over this period, any discrepancy in measurements can be attributable to observation error. The standard deviation of these discrepancies was estimated to be \Sexpr{obs_error_sigma}, meaning that 95\% of the measurement errors were within $\pm$ 1.47 cm of the true diameter. Having an estimate of observation error, we then used Bayesian point estimation to maximize the joint posterior of the \textit{``true''} initial and final diameter, and subsequently, \textit{``true growth''} for each individual (SUPP).

\subsection*{Estimating mortality rates}

\subsection*{Model fitting}

For all models we used Bayesian inference and fitted models in R \Sexpr{paste(version$major, version$minor, sep='.')} using package \texttt{rstan} \Sexpr{packageVersion("rstan")}\cite{StanDevelopmentTeam:2015uz}. We executed three independent chains and in all cases modelled parameters converged within 2000 iterations. Convergence was assessed through both visual inspection of chains and reference to the Brooks-Gelman-Rubin convergence diagnostic \cite{Gelman:1992ts, Brooks:1998ju}. After discarding the first 2000 iterations as ‘burn in’, a further 2000 iterations were taken from the joint posterior.
